---
title: "UN Data Exploration: An Exploration of the Tidyverse"
output: html_notebook
---

In this project, I'll be redoing the UN Data project using the tidyverse libraries. I'll be working with three csv files, gdp_per_capita.csv, life_expectancy.csv, and continents.csv.

```{r}
install.packages("corrr")
library(corrr)
library(tidyverse)
```

1. Using the read_csv() function, read the GDP dataset into your notebook as a tibble called gdp_df. After reading it in, inspect the first 10 rows and then inspect the last 10 rows.

```{r}
gdp_df <- read_csv('../data/gdp_per_capita.csv')
```

```{r}
gdp_df |> 
  slice_head(n=10)
```

```{r}
gdp_df |>
  slice_tail(n=10)
```

2. Drop the 'Value Footnotes' column, and rename the remaining columns to 'Country', 'Year', and 'GDP_Per_Capita'.

```{r}
gdp_df = gdp_df |>
  select(-'Value Footnotes') |>
  rename(Country='Country or Area', GDP_Per_Capita=Value)
gdp_df
```
We see there are no NA values for GDP per capita.
```{r}
gdp_df |>
  filter(is.na(GDP_Per_Capita))
```

3. Which years are represented in this dataset? How many observations are there per year? Make a plot to view the number of observations per year.

```{r}
start_year = gdp_df |>
  summarize(min_year = min(Year)) |>
  pull(min_year)
paste0('The first year in the dataset is ',start_year)

end_year = gdp_df |>
  summarize(max_year = max(Year)) |>
  pull(max_year)
paste0('The last year in the dataset is ',end_year)
```

```{r}
# Testing for any missing years
distinct_years = num_obs_per_year |>
  mutate(Year_Manual = start_year + row_number() - 1) |>
  mutate(Year_Match = Year == Year_Manual) |>
  summarize(year_match = sum(Year_Match)) |>
  pull(year_match)
distinct_years
```
We see that all 33 years from 1990 to 2022 are represented in the GDP dataset. 

```{r}
num_obs_per_year = gdp_df |>
  count(Year) |>
  arrange(Year) 
num_obs_per_year
```

```{r}
gdp_df |>
  count(Year) |>
  arrange(Year) |>
  ggplot(aes(x=Year, y=n)) +
  geom_line() + 
  labs(title = "Number of GDP Observations per Year",
    subtitle = "In the UN GDP Dataset",
    x = "Year", y = "Number of Observations")
```

4. How many countries are represented in this dataset? Which countries have the fewest observations?

```{r}
total_countries = gdp_df |>
  distinct(Country) |>
  count() |>
  pull()

paste0('There are ',total_countries,' represented in the GDP dataset.')
```

Countries with the fewest observations (by raw count):

```{r}
gdp_df |>
  count(Country) |>
  slice_min(n, n=3)
```

Countries in the bottom 10% of observations:
```{r}
gdp_df |>
  count(Country) |>
  mutate(Count_Percentile = percent_rank(n)) |>
  filter(Count_Percentile <= 0.10) |>
  arrange(Count_Percentile)
```


5. Create a new tibble by subsetting gdp_df to just the year 2021. Call this new tibble gdp_2021.

```{r}
gdp_2021 = gdp_df |>
  filter(Year == 2021)
```


6. Use summary() to find the summary statistics for GDP per capita in 2021.

```{r}
gdp_2021 |> 
  summarize(avg_gdp = mean(GDP_Per_Capita, na.rm = TRUE),
              median_gdp = median(GDP_Per_Capita, na.rm = TRUE),
              min_gdp = min(GDP_Per_Capita, na.rm = TRUE),
              max_gdp = max(GDP_Per_Capita, na.rm = TRUE),
              sd_gdp = sd(GDP_Per_Capita, na.rm = TRUE)
  )
```

7. Create a histogram of GDP Per Capita numbers for 2021 (you may wish to adjust the number of bins for your histogram).

```{r}
gdp_df |>
  ggplot(aes(x=GDP_Per_Capita)) +
  geom_histogram(bins = 20) + 
  labs(title = "National GDP Per Capita Values in 2021",
    subtitle = "In the UN GDP Dataset",
    x = "GDP Per Capita (in PPP)", y = "Count")
```

8. Find the top 5 countries and bottom 5 countries by GDP per capita in 2021.

Top 5 countries by GDP per capita in 2021:
```{r}
gdp_2021 |>
  slice_max(GDP_Per_Capita, n=5)
```
Bottom 5 countries by GDP per capita in 2021:
```{r}
gdp_2021 |>
  slice_min(GDP_Per_Capita, n=5)
```

9. Now, return to the full dataset, gdp_df. Pivot the data for 1990 and 2021 (using pivot_wider() function) so that each row corresponds to a country, each column corresponds to a year, and the values in the table give the GDP_Per_Capita amount. Drop any rows that are missing values for either 1990 or 2021. Save the result to a tibble named gdp_pivoted.

```{r}
gdp_pivoted = gdp_df |>
  filter(Year==1990 | Year==2021) |>
  pivot_wider(id_cols=Country, names_from=Year, values_from=GDP_Per_Capita) |>
  drop_na()
```


10. Create a new column in gdp_pivoted named Percent_Change. This column should contain the percent change in GDP_Per_Capita from 1990 to 2021. Hint: Percent change is calculated as 100*(New Value - Old Value) / Old Value.

```{r}
gdp_pivoted = gdp_pivoted |>
  mutate(Percent_Change=100*(`2021` - `1990`) / `1990`)
```


11. How many countries experienced a negative percent change in GDP per capita from 1990 to 2021?

```{r}
neg_change = gdp_pivoted |>
  filter(Percent_Change < 0) |>
  count() |>
  pull()

paste0('There were ',neg_change,' countries that experienced a negative percent change in GDP per capita from 1990 to 2021.')
```


12. Which country had the highest % change in GDP per capita? Which country had the second highest % change? Create a line plot showing these country's GDP per capita for all years for which you have data. Put both line charts on the same plot.

```{r}
top_2_gdp_increase = gdp_pivoted |>
  slice_max(Percent_Change, n=2) |>
  pull(Country)

paste0(top_2_gdp_increase[1],' had the highest percent increase in GDP per capita from 1990 to 2021.')
paste0(top_2_gdp_increase[2],' had the second highest percent increase in GDP per capita from 1990 to 2021.')

```

```{r}
gdp_df |>
  filter(Country %in% top_2_gdp_increase) |>
  ggplot(aes(x=Year, y=GDP_Per_Capita)) +
  geom_line(aes(color=Country)) + 
  labs(title = "GDP Per Capita, 1990 - 2021",
  subtitle = "Top 2 Countries with Highest Percent Increase in GDP",
  x = "Year", y = "GDP Per Capita (in PPP)")
```


13. Read in continents.csv contained in the data folder into a new tibble called continents. We will be using this tibble to add a new column to our dataset.

```{r}
continents <- read_csv('../data/continents.csv')
```


14. Merge gdp_df and continents. Keep only the countries that appear in both data frames. Save the result back to gdp_df.

```{r}
gdp_df = inner_join(gdp_df, continents, by='Country')
```


15. Determine the number of countries per continent. Create a bar chart showing this.

Number of countries per continent:
```{r}
gdp_df |>
  group_by(Continent) |>
  distinct(Country) |>
  count()
```
Plot showing number of countries per continent:
```{r}
gdp_df |>
  group_by(Continent) |>
  distinct(Country) |>
  ggplot(aes(x=Continent)) +
  geom_bar() +
  labs(title = "Number of Countries per Continent",
  x = "Continent", y = "Country Count")
```

16. Create a boxplot showing GDP per capita in 2021 split out by continent. What do you notice?

```{r}
gdp_df = gdp_df |>
  mutate(GDP_Per_Capita_Thousands = GDP_Per_Capita/1000)

gdp_df |>
  ggplot(aes(x=Continent, y=GDP_Per_Capita_Thousands)) +
  geom_boxplot() +
  labs(title = "GDP Distribution by Continent, 2021",
  x = "Continent", y = "GDP Per Capita (thousands PPP)")
```
In 2021, the GDP per capita in African, Oceanian, and South American countries tends to be low (50,000 PPP or lower). The top 5 highest-GDP countries in 2021 are all in Asia. Europe has the highest proportion of countries with a GDP of 30,000+ PPP (about 50%). All continents have at least 2-3 extreme outliers, representing countries with a very high GDP compared with other countries in the same continent. The GDP distribution for all continents is right-skewed, with the skew being more pronounced for some continents (Africa, Oceania, and Asia) and less pronounced for others (Europe, North America, and South America).

17. Read life_expectancy.csv into a tibble named life_expectancy. Do not modify the csv file in order to read this data in.

```{r}
life_expectancy <- read_csv('../data/life_expectancy.csv', skip=3)
```
18. Drop the Country Code, Indicator Name, and Indicator Code columns. Then use pivot_longer to convert your data from wide to long. That is, instead of having one row per country and multiple columns per year, we want to have multiple rows per country and a single column for year. After melting, rename the columns to Country, Year, and Life_Expectancy.

```{r}
life_expectancy = life_expectancy |>
  select(-'Country Code',-'Indicator Name',-'Indicator Code') |>
  pivot_longer(cols=2:last_col(), names_to='Year', values_to='Life_Expectancy') |>
  rename(Country='Country Name')
life_expectancy |> head(2)
```

19. What was the first country with a life expectancy to exceed 80?

```{r}
# Converting life_expectancy$Year to a numeric datatype since we will be finding the minimum year
life_expectancy = life_expectancy |>
  mutate(Year = as.double(Year)) |>
  drop_na(Year)
```

Japan was the first country with a life expectancy to exceed 80:
```{r}
life_expectancy |> 
  filter(Life_Expectancy > 80) |>
  slice_min(Year) |> 
  pull(Country)
```

20. Merge gdp_df and life_expectancy, keeping all countries and years that appear in both tibbles. Save the result to a new tibble named gdp_le. If you get any errors in doing this, read them carefully and correct them. Look at the first five rows of your new data frame to confirm it merged correctly. Also, check the last five rows to make sure the data is clean and as expected.

```{r}
gdp_le = inner_join(gdp_df, life_expectancy, by=c('Country','Year'))
```

```{r}
gdp_le |> head(5)
```

```{r}
gdp_le |> tail(5)
```

21. Create a new tibble, named gdp_le_2021 by extracting data for the year 2021 from gdp_le. How many countries have a life expectancy of at least 80 in 2021?

```{r}
gdp_le_2021 = gdp_le |>
  filter(Year==2021)
gdp_le_2021 |> head(2)
```

22. Find the countries that had the top 3 largest GDP per capita figures for 2021. Create a plot showing the change in life expectancy over time for these three countries. This plot should be faceted so that each country is contained in its own figure.

```{r}
top_3_gdp_2021 = gdp_le_2021 |>
  slice_max(GDP_Per_Capita, n=3) |>
  pull(Country)

paste0('Countries with the top 3 largest GDPs in 2021: ')
top_3_gdp_2021
```

```{r}
top_3_gdp = gdp_le |>
  filter(Country %in% top_3_gdp_2021)
  
ggplot(top_3_gdp, aes(x = Year, y = Life_Expectancy)) +
  geom_line() +
  facet_wrap(~Country) + 
  labs(title = "GDP Per Capita, 1990 - 2022",
  x = "Year", y = "Life Expectancy (in years)")
```


23. Create a scatter plot of Life Expectancy vs GDP per Capita for the year 2021. What do you notice?

```{r}
ggplot(gdp_le_2021, aes(x = GDP_Per_Capita_Thousands, y = Life_Expectancy)) +
  geom_point() +
  labs(title = "Life Expectancy vs GDP per Capita, 2021",
  x = "GDP Per Capita (thousands PPP)", y = "Life Expectancy (in years)")
```
The scatterplot of Life Expectancy vs GDP per Capita for the year 2021 shows a curved relationship, where average life expectancy rises rapidly as GDP increases from 0 to 30,000 PPP, then increases at a much slower rate as GDP per capita increases from 30,000 to 120,000 PPP.

24. Find the correlation between Life Expectancy and GDP per Capita for the year 2021. What is the meaning of this number?

```{r}
le_gdp_cor <- gdp_le_2021 |>
  select(Life_Expectancy, GDP_Per_Capita) |>
  correlate()

le_gdp_cor
```
The correlation between Life Expectancy and GDP per Capita for the year 2021 is about 0.745. This indicates a strong positive linear relationship between life expectancy and GDP per capita; that is, an increase in GDP per capita tends to be associated with an increase in life expectancy. 

25. Add a column to gdp_le_2021 and calculate the logarithm of GDP per capita. Find the correlation between the log of GDP per capita and life expectancy. How does this compare to the calculation in the previous part? Look at a scatter plot to see if the result of this calculation makes sense.

```{r}
gdp_le_2021 = gdp_le_2021 |>
  mutate(Log_GDP_Per_Capita=log(GDP_Per_Capita))

le_log_gdp_cor <- gdp_le_2021 |>
  select(Life_Expectancy, Log_GDP_Per_Capita) |>
  correlate()

le_log_gdp_cor
```
Taking the log of GDP per capita increases the correlation coefficient by about 0.1, from 0.745 to 0.847. Thus we have a very strong positive correlation between Life Expectancy and log GDP per Capita for the year 2021.

```{r}
ggplot(gdp_le_2021, aes(x = Log_GDP_Per_Capita, y = Life_Expectancy)) +
  geom_point() +
  labs(title = "Life Expectancy vs Log GDP per Capita, 2021",
  x = "Log GDP Per Capita (thousands PPP)", y = "Life Expectancy (in years)")
```
The increased correlation coefficient makes sense when comparing the scatterplots. The scatterplot showing the relationship between GDP per capita and life expectancy resembles a curve; while the scatterplot of life expectancy vs log GDP per capita shows a linear relationship. 








